Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-06-23 paper.qmd
Update low-density bins

  bin2 <- calc_bins_y(bin1 = xbins, r2 = r2)$bin2

  scurve_model <- fit_highd_model(
    training_data = training_data_scurve,
    emb_df = umap_scurve_scaled,
    bin1 = xbins,
    r2 = r2,
    is_bin_centroid = TRUE,
    is_rm_lwd_hex = FALSE,
    col_start_highd = "x"
  )

@@

```{r}
## To initialize effective bins along x
effective_bin1_scurve <- 10

effective_bin2_scurve <- calc_bins_y(bin1 = effective_bin1_scurve, r2 = r2)$bin2

scurve_model <- fit_highd_model(
  training_data = training_data_scurve,
  emb_df = umap_scurve_scaled,
  bin1 = effective_bin1_scurve,
  r2 = r2,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "x"
)

@@

```{r}
cell_count_scurve <- ggplot(df_bin_centroids_scurve, 
                            aes(x = reorder(as.factor(hexID), 
                                            -std_counts), 
                                y = std_counts)) + 
  geom_quasirandom() + 
  geom_hline(yintercept = benchmark1, color = "black", linewidth=0.8, alpha = 0.5) +
  xlab("hexagon id (re-ordered)") + 
  ylab("standardized bin count") +
  ggtitle("(b)") +
  theme(aspect.ratio = 0.75,
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10, angle = 90),
        #axis.text.x = element_text(size = 7, angle = 90),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 10),
        title = element_text(size = 36))
```

@@

scurve_model <- fit_highd_model(
  training_data = training_data_scurve,
  emb_df = umap_scurve_scaled,
  bin1 = effective_bin1_scurve,
  r2 = r2,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "x"
)

@@

<!--Using first quantile as the benchmark value-->
```{r}
hex_grid_with_low_counts <- full_join(hex_grid2, 
    df_bin_centroids_scurve, 
    by = c("hex_poly_id" = "hexID")) |>
  filter(!is.na(std_counts)) |>
  mutate(type = if_else(std_counts <= benchmark1, "low", "high")) |>
  filter(type == "low")

hex_grid_rm_lwd_scurve1 <-  ggplot(
  data = hex_grid_with_counts_s_curve2, 
  aes(x = x, y = y)) + 
  geom_polygon(aes(group = hex_poly_id, fill = std_counts), 
            color = "grey70", linewidth=0.2) +
  geom_text(data = hex_grid_with_low_counts, 
            color = "black", 
            aes(x = c_x, y = c_y, group = hex_poly_id, label = "X", size = 1)) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("a")
```

<!--Observing neighboring bins as well-->
```{r}
## First define low density bins using first quantile
df_bin_centroids_low <- df_bin_centroids_scurve |> 
  filter(std_counts <= benchmark1)

## Check neighboring bins
remove_id <- find_low_dens_hex(df_bin_centroids_all = df_bin_centroids_scurve, 
                               bin1 = effective_bin1_scurve, 
                               df_bin_centroids_low = df_bin_centroids_low)

## Remove the identified bins
df_bin_centroids_scurve_removed <- df_bin_centroids_scurve |>
  filter(hexID %in%remove_id)

df_bin_centroids_scurve_keep  <- df_bin_centroids_scurve |>
  filter(!(hexID %in%remove_id))

@@

hex_grid_rm_lwd_scurve2 <-  ggplot(
  data = hex_grid_with_counts_s_curve2, 
  aes(x = x, y = y)) + 
  geom_polygon(aes(group = hex_poly_id, fill = std_counts), 
            color = "grey70", linewidth=0.2) +
  geom_text(data = hex_grid_with_low_counts_neighbor, 
            color = "black", 
            aes(x = c_x, y = c_y, group = hex_poly_id, label = "X", size = 1)) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b")
```

>>>>

  bin2 <- calc_bins_y(bin1 = xbins, r2 = r2, q = 0.07)$bin2

  scurve_model <- fit_highd_model(
    training_data = training_data_scurve,
    emb_df = umap_scurve_scaled,
    bin1 = xbins,
    r2 = r2,
    q = 0.07, 
    is_bin_centroid = TRUE,
    is_rm_lwd_hex = FALSE,
    col_start_highd = "x"
  )

@@

```{r}
## To initialize effective bins along x
effective_bin1_scurve <- 12

effective_bin2_scurve <- calc_bins_y(
  bin1 = effective_bin1_scurve, 
  r2 = r2, 
  q = 0.07)$bin2

scurve_model <- fit_highd_model(
  training_data = training_data_scurve,
  emb_df = umap_scurve_scaled,
  bin1 = effective_bin1_scurve,
  r2 = r2,
  q = 0.07,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "x"
)

@@

```{r}
cell_count_scurve <- ggplot(df_bin_centroids_scurve, 
                            aes(x = reorder(as.factor(hexID), 
                                            -std_counts), 
                                y = std_counts)) + 
  geom_quasirandom() + 
  geom_hline(yintercept = benchmark1, color = "black", linewidth=0.8, alpha = 0.5) +
  xlab("hexagon id (re-ordered)") + 
  ylab("standardized bin count") +
  ggtitle("(b)") +
  theme(aspect.ratio = 0.75,
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10, angle = 90),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 10),
        title = element_text(size = 36))
```

@@

scurve_model <- fit_highd_model(
  training_data = training_data_scurve,
  emb_df = umap_scurve_scaled,
  bin1 = effective_bin1_scurve,
  r2 = r2,
  q = 0.07, 
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "x"
)

@@

<!--Using first quantile as the benchmark value-->
```{r}
## hexagon binning to have regular hexagons
hb_obj_scurve4 <- hex_binning(
  data = umap_scurve_scaled, 
  bin1 = 12, 
  r2 = r2,
  q = 0.07)

## Data set with all centroids
all_centroids_df4 <- hb_obj_scurve4$centroids

## Generate all coordinates of hexagons
hex_grid4 <- hb_obj_scurve4$hex_poly

## To obtain the standardise counts within hexbins
counts_df4 <- hb_obj_scurve4$std_cts
df_bin_centroids4 <- extract_hexbin_centroids(
  centroids_df = all_centroids_df4, 
  counts_df = counts_df4) |> 
  filter(drop_empty == FALSE)

hex_grid_with_counts_s_curve4 <- full_join(
  hex_grid4, 
  df_bin_centroids4 |> select(hexID, std_counts), 
  by = c("hex_poly_id" = "hexID"))

hex_grid_with_low_counts <- full_join(hex_grid4, 
    df_bin_centroids_scurve, 
    by = c("hex_poly_id" = "hexID")) |>
  filter(!is.na(std_counts)) |>
  mutate(type = if_else(std_counts <= benchmark1, "low", "high")) |>
  filter(type == "low")

hex_grid_rm_lwd_scurve1 <-  ggplot(
  data = hex_grid_with_counts_s_curve4, 
  aes(x = x, y = y)) + 
  geom_polygon(aes(group = hex_poly_id, fill = std_counts), 
            color = "grey70", linewidth=0.2) +
  geom_text(data = hex_grid_with_low_counts, 
            color = "black", 
            aes(x = c_x, y = c_y, group = hex_poly_id, label = "X", size = 1)) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("a")
```

<!--Observing neighboring bins as well-->
```{r}
## First define low density bins using first quantile
df_bin_centroids_low <- df_bin_centroids4 |> 
  filter(std_counts <= benchmark1)

## Check neighboring bins
remove_id <- find_low_dens_hex(df_bin_centroids_all = df_bin_centroids4, 
                               bin1 = effective_bin1_scurve, 
                               df_bin_centroids_low = df_bin_centroids_low)

## Remove the identified bins
df_bin_centroids_scurve_removed <- df_bin_centroids4 |>
  filter(hexID %in%remove_id)

df_bin_centroids_scurve_keep  <- df_bin_centroids4 |>
  filter(!(hexID %in%remove_id))

@@

hex_grid_rm_lwd_scurve2 <-  ggplot(
  data = hex_grid_with_counts_s_curve4, 
  aes(x = x, y = y)) + 
  geom_polygon(aes(group = hex_poly_id, fill = std_counts), 
            color = "grey70", linewidth=0.2) +
  geom_text(data = hex_grid_with_low_counts_neighbor, 
            color = "black", 
            aes(x = c_x, y = c_y, group = hex_poly_id, label = "X", size = 1)) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b")
```
